---
title: "Machine Learning"
subtitle: "A hands-on introduction using R"
author: "Sean Davis"
date: "July 7, 2017"
output: 
  revealjs::revealjs_presentation:
    widescreen: true
    keep_md: true
    theme: solarized
---

<style type="text/css">
  .reveal p {
    text-align: left;
  }
  .reveal ul {
    display: block;
  }
  .reveal ol {
    display: block;
  }  
</style>

# Preliminaries

## Install required libraries.

```{r echo=FALSE,results='hide'}
library(knitr)
options(width=60)
opts_chunk$set(tidy=TRUE,warning=FALSE,message=FALSE, cache=TRUE,
               fig.width=9,fig.height=5.5)
```

```{r eval=FALSE}
library(BiocInstaller)
biocLite(c('mlbench','adabag', 'e1071', 'randomForest', 'party', 'mboost', 'rpart.plot', 'formatR'))
```

```{r}
require(c('mlbench','adabag', 'e1071', 'randomForest', 'party', 'mboost', 'rpart.plot', 'formatR'))
```

### Some links of interest

- [caret](https://https://cran.r-project.org/package=caret), [party](https://cran.r-project.org/package=party), [randomForest](https://cran.r-project.org/package=randomForest), [mlbench](https://cran.r-project.org/package=mlbench), [mlr](https://cran.r-project.org/package=mlr)
- [Max Kuhn's old machine learning tutorial](https://www.r-project.org/conferences/useR-2013/Tutorials/kuhn/user_caret_2up.pdf)

# Overview

## What is machine learning?

Machine learning is a broad set of fields related to computers learning from "experience" (data). 

- Focusing on *predictive modeling* with a goal of *producing the most accurate estimates of some quantity or the most likely output of an event*. 
- These models are sometimes based on similar models for inference (testing against a null hypothesis, such as linear regression), but in many cases, predictive models are not well-suited for inference (think k-nearest-neighbor, for example). 

## The formula interface

```{r eval=FALSE}
outcome ~ var1 + var2 + ...
```

The variable `outcome` is predicted by `var1, var2, ...`

```{r eval=FALSE}
some_model_function(price ~ numBedrooms + numBaths + acres,
                 data = housingData)
```

Conveniences of the formula interface:

- Transformations such as `log10(acres)` can be specified inline. 
- Factors are converted into dummy variables automatically.

## The Non-formula interface

- The non-formula interface specifies the predictors as a matrix or data frame. 
- The outcome data are then passed into the model as a vector.

```{r eval=FALSE}
some_model_function(x = housePredictors, y = price)
```

Many R functions offer both a formula and a non-formula interface, but not all.

## General workflow for machine learning in R

1. Fit the model to a set of training data
    ```{r eval=FALSE}
    fit <- knn(trainingData, outcome, k = 5)
    ```
2. Assess the properties of the model using `print`, `plot`, `summary` or other methods
3. Predict outcomes for samples using the predict method:
    ```{r eval=FALSE}
    predict(fit, newSamples).
    ```

# Exercise 1
Playing with regression

## Is `mpg` a function of `wt`? {.smaller}

The formula interface in action:

```{r}
data(mtcars)
fit = lm(mpg ~ wt, data = mtcars)
summary(fit)
```

And make a plot.

```{r eval=FALSE}
plot(mpg ~ wt, data=mtcars)
abline(fit)
```

## Is `mpg` a function of `wt`?

```{r echo=FALSE}
plot(mpg ~ wt, data=mtcars)
abline(fit)
```

## Use `wt` to predict `mpg`

And predict the original data based on the fitted model.

```{r}
pred_mpg = predict(fit,mtcars)
summary(pred_mpg)
```

And look at the predicted values:

```{R eval=FALSE}
plot(mpg ~ wt, data=mtcars)
abline(fit)
points(y = pred_mpg, x = mtcars$wt, col='red')
```

## Use `wt` to predict `mpg`

```{R echo=FALSE}
plot(mpg ~ wt, data=mtcars)
abline(fit)
points(y = pred_mpg, x = mtcars$wt, col='red')
```

## Quantifying "goodness-of-fit"

```{R echo=FALSE}
plot(mpg ~ wt, data=mtcars)
abline(fit)
points(y = pred_mpg, x = mtcars$wt, col='red')
```

## Quantifying "goodness-of-fit"

- Residual Sum of Squares
    $$ RSS = \sum_{N} (y_i - f(x_i))^{2} $$

## Quantifying "goodness-of-fit"

```{r}
rss = sum((mtcars$mpg - predict(fit,mtcars))^2)
rss
anova(fit)
```

## Training versus testing

- What did we do wrong in quantifying our "goodness-of-fit"?

# Splitting data and model performance evaluation


## Common steps during training

- estimating model parameters (i.e. training models)
- determining the values of tuning parameters that cannot be directly calculated from the data
- calculating the performance of the final model that will generalize to new data

## Spending the data to find an optimal model? 

- Split data into training and test data sets
- *Training Set*: these data are used to estimate model parameters and
to pick the values of the complexity parameter(s) for the model.
- *Test Set (aka validation set)*: these data can be used to get an independent assessment of model accuracy. The test data should never be used in any aspect of model training.

## Tradeoffs in spending data

The more data we spend, the better estimates we’ll get (provided the data is accurate). Given a fixed amount of dat:

- Too much spent in training won’t allow us to get a good assessment of predictive performance. We may find a model that fits the training data very well, but is not generalizable (over–fitting)
- Too much spent in testing won’t allow us to get a good assessment of
model parameters


## Example using `mtcars`

Using 50% of the data for training and 50% for testing is a place to start.

```{r}
set.seed(1)
trainIdx = sample(1:nrow(mtcars),16)
trainDat = mtcars[trainIdx,]
testDat  = mtcars[-trainIdx,]
```

## Train the model using the training data

```{r}
fit = lm(mpg ~ wt, data=trainDat)
anova(fit)
```

## Train the model using the training data

```{r echo=FALSE}
plot(mpg ~ wt, data = trainDat)
abline(fit)
```

## Test our model using the testing data

```{r}
pred_mpg = predict(fit,testDat)
rss = sum((testDat$mpg - pred_mpg)^2)
rss
anova(fit)
```

## Test our model using the testing data

```{r echo=FALSE}
plot(mpg ~ wt, data = testDat)
points(testDat$wt,pred_mpg,col='red')
abline(fit)
```

# Example 2
A classification example

## Classification Trees

As a simple dataset to try with machine learning, we are going to predict the species of 
`iris` based on four measurements.

```{r eval=FALSE}
data(iris)
View(iris)
pairs(iris[,1:4],col=iris$Species)
```

## Iris Data

```{r echo=FALSE}
data(iris)
pairs(iris[,1:4],col=iris$Species)
```


## Another slide

We can start with a simple learner, a [classification tree](https://en.wikipedia.org/wiki/Decision_tree_learning). This learner requires:

- A known class for each observation
- A set of "features" that will serve a potential predictors

1. Start with whole dataset.
2. Choose features one-at-a-time and look for a value of each variable that ends up with the most homogeneous two groups after splitting on that variable/value.
3. For each resulting group, repeat step 2 until all remaining groups have only one class in them.
4. Optionally, "prune" the tree to keep only splits that are "statistically significant".

## Learning the model

The `party` package includes a function, `ctree` to "learn" a tree from data.

```{r eval=FALSE}
library(party)
x = ctree(Species ~ .,data=iris)
plot(x)
```

## Learning the model

```{r echo=FALSE}
library(party)
x = ctree(Species ~ .,data=iris)
plot(x)
```



## Checking the model

```{r}
library(caret)
library(e1071)
prediction = predict(x,iris)
table(prediction)
confusionMatrix(iris$Species,prediction)
```

## Data splitting, take 2

What is the problem with what we just did to determine our prediction accuracy?  
To deal with this problem, we can split the dataset into a "training" set and then check
our prediction on the other piece of the data, the "test" set.

```{r}
set.seed(42)
trainIdx = sample(c(TRUE,FALSE),size=nrow(iris),prob=c(0.2,0.8),replace=TRUE)
irisTrain = iris[trainIdx,]
irisTest  = iris[!trainIdx,]
nrow(irisTrain)
nrow(irisTest)
```

## "train" our tree on the "training" set.

```{r eval=TRUE}
trainTree = ctree(Species ~ ., data = irisTrain)
plot(trainTree)
```

## "train" our tree on the "training" set.

```{r echo=FALSE}
trainTree = ctree(Species ~ ., data = irisTrain)
plot(trainTree)
```

## Test our predictions on the "training" data

And how does our `trainTree` do at predicting the original classes in the "training" data?

```{r eval=FALSE}
trainPred = predict(trainTree,irisTrain)
confusionMatrix(irisTrain$Species,trainPred)
```

##  Test our predictions on the "training" data

```{r echo=FALSE}
trainPred = predict(trainTree,irisTrain)
confusionMatrix(irisTrain$Species,trainPred)
```

## Test our predictions on the "testing" data

How is our prediction performance now on the "test" data?

```{r eval=FALSE}
testPred = predict(trainTree,irisTest)
confusionMatrix(irisTest$Species,testPred)
```

## Test our predictions on the "testing" data

```{r echo=FALSE}
testPred = predict(trainTree,irisTest)
confusionMatrix(irisTest$Species,testPred)
```

# Example 3
k-nearest neighbor and cross-validation

## k-nearest-neighbor

Now, let's make this harder. We will now look at a dataset that is designed to "foil" classifiers. 

```{r eval=FALSE}
library(mlbench)
set.seed(1)
spiral = mlbench.spirals(1000,sd=0.1)
spiral = data.frame(x=spiral$x[,1],y=spiral$x[,2],class=factor(spiral$classes))
library(ggplot2)
ggplot(spiral,aes(x,y,color=class)) + geom_point()
```

## k-nearest-neighbor

```{r echo=FALSE}
library(mlbench)
set.seed(1)
spiral = mlbench.spirals(1000,sd=0.1)
spiral = data.frame(x=spiral$x[,1],y=spiral$x[,2],class=factor(spiral$classes))
library(ggplot2)
ggplot(spiral,aes(x,y,color=class)) + geom_point()
```

## Without splitting data

```{r}
library(caret)
fit = knn3(class ~ ., data=spiral)
confusionMatrix(predict(fit,spiral,type='class'),spiral$class)
```

## Cross-validation

setup

```{r}
library(caret)
indxTrain <- createDataPartition(y = spiral$class,p = 0.75,list = FALSE)
training <- spiral[indxTrain,]
testing <- spiral[-indxTrain,]
ctrl <- trainControl(method="repeatedcv",repeats = 3)
knnFit <- train(class ~ ., data = training, method = "knn", trControl = ctrl, tuneLength = 10)
```

## Cross-validation {.smaller}

```{r echo=FALSE}
knnFit
```

# Exercise 4
Ensembles of learners

## What is an ensemble of learners?

In some cases, a machine learning algorithm can have limited predictive power, but using multiple "instances" of such *weak learners* in combination can produce a good result.

It is probably obvious that a classification tree approach might be problematic for a dataset like the `spiral` dataset. In this example, we are going to use "boosting" to combine many trees, each with minimal prediction capabilities, into an "ensemble" learner with reasonable good prediction capabilities.

## Using trees to predict on the `spiral` dataset

```{r}
library(formatR)
library(party)
trainIdx = sample(c(TRUE,FALSE),nrow(spiral),replace=TRUE,prob=c(0.5,0.5))
spiralTrain = spiral[trainIdx,]
trainTree   = ctree(class ~ .,spiralTrain)
```

## Using trees to predict on the `spiral` dataset


```{r echo=FALSE}
plot(trainTree)
```

## Using trees to predict on the `spiral` dataset {.smaller}

Training Data

```{r}
prediction = predict(trainTree,spiralTrain)
confusionMatrix(spiralTrain$class,prediction)
```

## Using trees to predict on the `spiral` dataset {.smaller}

Testing data

```{r}
spiralTest = spiral[!trainIdx,]
prediction = predict(trainTree,spiralTest)
confusionMatrix(spiralTest$class,prediction)
```

## Using trees to predict on the `spiral` dataset 

Many trees have similar prediction capability, but each is really bad.  This is a 
characteristic of a "weak learner".  Here, we see that in action by performing a bootstrap
sampling (resample with replacement), train, plot, and check prediction accuracy.

## Using trees to predict on the `spiral` dataset

Must be run "locally" to see effect.

```{r}
plotBootSample = function(spiral) {
  trainIdx = sample(1:nrow(spiral),replace=TRUE)
  spiralTrain = spiral[trainIdx,]
  trainTree   = ctree(class ~ .,spiralTrain)
  plot(trainTree)
  spiralTest = spiral[-trainIdx,]
  prediction = predict(trainTree,spiralTest)
  print(confusionMatrix(spiralTest$class,prediction)$overall['Accuracy'])
}
```

```{r eval=FALSE}
# press 'ESC' or 'ctrl-c' to stop
while(TRUE) {
  par(ask=TRUE)
  plotBootSample(spiral)
}
```

## Boosting

We can "combine" a bunch of "weak learners", giving more "weight" to hard-to-classify observations as we build each new classifier.  In this case, we will be using the same classification tree approach again.

```{r}
library(adabag)
trainIdx      = sample(c(TRUE,FALSE),nrow(spiral),replace=TRUE,prob=c(0.5,0.5))
spiralTrain   = spiral[trainIdx,]
boostTree     = boosting(class ~ x + y,data = spiralTrain,control = rpart.control(maxdepth=2))
prediction    = predict(boostTree,spiralTrain)
```

## Boosting results {.smaller}

```{r echo=FALSE}
confusionMatrix(spiralTrain$class,prediction$class)
```

## A few trees from our ensemble

```{r echo=FALSE}
library(rpart.plot)
par(mfrow=c(2,3),ask=FALSE)
for(i in 1:6) {
  rpart.plot(boostTree$trees[[i]])
}
```

## Boosted trees on test data

```{r eval=FALSE}
spiralTest = spiral[!trainIdx,]
prediction = predict(boostTree,spiralTest)
confusionMatrix(spiralTest$class,prediction$class)
```

## Boosted trees on test data

```{r echo=FALSE}
spiralTest = spiral[!trainIdx,]
prediction = predict(boostTree,spiralTest)
confusionMatrix(spiralTest$class,prediction$class)
```


# Exercise 5
Random forests--ensembles to the max

## Random Forests

```{r}
library(randomForest)
res = randomForest(Species ~ .,data=iris)
res
```

# sessionInfo

## sessionInfo

```{r echo=FALSE}
sessionInfo()
```


    <script>
        (function(i,s,o,g,r,a,m){i['GoogleAnalyticsObject']=r;i[r]=i[r]||function(){
        (i[r].q=i[r].q||[]).push(arguments)},i[r].l=1*new Date();a=s.createElement(o),
        m=s.getElementsByTagName(o)[0];a.async=1;a.src=g;m.parentNode.insertBefore(a,m)
        })(window,document,'script','https://www.google-analytics.com/analytics.js','ga');
        ga('create', 'UA-93043521-1', 'auto');
        ga('send', 'pageview');

         
        var links = document.querySelectorAll('a');
        Array.prototype.map.call(links, function(item) {
            if (item.host != document.location.host) {
                item.addEventListener('click', function() {
                    var action = item.getAttribute('data-action') || 'follow';
                    ga('send', 'event', 'outbound', action, item.href);
                });
            }
        });
    </script>
    